Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-02-02 reviews/response-editor.tex

\begin{itemize}

    \item \textbf{Sect.~2.1 (Model overview):}
    We have clarified that ETOPO1 data were bilinearly interpolated to the
    model grids.

    \item \textbf{Sect.~2.1 (Model overview):}
    We have explained the new subdivisions of Sect.~2 (Model setup).

    \item \textbf{Sect.~2.2 (Ice rheology):}
    In this new subsection, we have added much detail on the model physics and
    on our choice of parameter for the sensitivity study to ice rheological
    parameters.

    \item \textbf{Sect.~2.2 (Ice rheology):}
    We removed the formulation ``SSA as a sliding law for the SIA'' which could
    induce in error.

@@

    \item \textbf{Sect.~2.2 (Ice rheology):}
    We have made clear that temperatures are computed in the bedrock to a depth
    of 3\,km below the ice sheet base where geothermal heat flux is applied.

    \item \textbf{Sect.~2.3 (Basal sliding):}
    In this new subsection, we have added much detail on the model physics and
    on our choice of parameter for the sensitivity study to basal sliding
    parameters.

@@

    \item \textbf{Sect.~2.6 (Climate forcing):}
    We have clarified that NARR temperature and precipitation fields were
    bilinearly interpolated to the model grids and admitted their limitations
    in reproducing steep climatic gradients characteristic of the study area.

    \item \textbf{Sect.~4 (Sensitivity to ice flow parameters):}
    In this new section, we present results of the new sensitivity study to ice
    rheological and basal sliding parameters

@@

    \item \textbf{Sect.~5.1.2 (Ice configuration during MIS 2):}
    We have recognized the limitations of the NARR with simulating orographic
    processes in some areas of steep topography and acknowledged that using
    downscaling methods could indeed reduce some of the inconsistence between
    model results and geological reconstructions.

    \item \textbf{Sect.~5.1.2 (Ice configuration during MIS 2):}
    We have balanced the final statement concerning modelled ice volume maxima
    using the uncertainty estimates obtained in the new Sect.~4 (Sensitivity to
    ice flow parameters).

    \item \textbf{Sect.~5.1.3 (Ice configuration during MIS 4):}
    We have balanced the final statement concerning modelled ice volume maxima
    using the uncertainty estimates obtained in the new Sect.~4 (Sensitivity to
    ice flow parameters).

    \item \textbf{Sect.~5.2.2 (Major ice-dispersal centres):}
    In the light of modelled accumulation areas, we have shown that present-day
    ice volumes contained in the ETOPO1 basal topography data likely does not
    affect the conclusions of our study.

    \item \textbf{Sect.~5.2.2 (Major ice-dispersal centres):}
    We have balanced the final statement concerning modelled ice volume minima
    using the uncertainty estimates obtained in the new Sect.~4 (Sensitivity to
    ice flow parameters).

    \item \textbf{Table~1:}
    The default parameter table was extended to include more parameter values
    and references to their source.

@@

    \item \textbf{Table~3:}
    This new table contains ice rheological and basal sliding parameter values
    used in the sensitivity study.

    \item \textbf{Table~5:}
    This new table contains modelled ice volume and area extremes corresponding
    to the results of the sensitivity study (similarly to Table~4).

@@

    \item \textbf{Fig.~2:}
    This new figure illustrates the different parametrisations of ice rheology
    used in the sensitivity study.

    \item \textbf{Fig.~3:}
    This new figure illustrates the different parametrisations of basal sliding
    used in the sensitivity study.

    \item \textbf{Fig.~7:}
    This new figure contains modelled sea-level relevant ice volume time series
    resulting from the sensitivity study (similarly to Fig.~5).

@@

    \item \textbf{Abstract:}
    We have reworked the first two sentences to clarify that the Cordilleran
    ice sheet is less understood than its counterparts not by lack of previous
    studies but due to its complexity.

    \item \textbf{Fig.~1:}
    We have enlarged the location map to include full geographic names instead
    of abbreviations.

    \item \textbf{Figs.~4, 6, 8--12, 14--16:}
    We have added rivers, lakes and graticules on all model output maps.

    \item \textbf{Sect.~5.2.3 (Erosional imprint on the landscape):}
    We have added a reference to Stumpf et~al. (2000).

    \item \textbf{Sect.~5.3.3 (Deglacial flow directions):}
    We have added references to Stumpf et~al. (2000) and to Ferbey et~al. (2013).

### 2016-02-02 cordillera-cycle.tex
Add even more corrections from Irina.

      Despite large differences in the input climate forcing
      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
      features that can be observed across the range of forcing data used.
      In all simulations, modelled sea-level relevant ice volumes remain
      relatively low during
      most of the glacial cycle, except during two major glacial events
      which occur between 62.2 and 56.9\,\unit{ka} during MIS~4, and between
      23.2 and 16.9\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
      panel). A sea-level relevant ice volume minimum is consistently
      reached between 55.9 and
      42.9\,\unit{ka} during MIS~3. However, the magnitude and precise
      timing of these three events depend significantly on the choice of
      proxy record used to derive a~time-dependent climate forcing
      (Table~\ref{tab:extrema}).

      Simulations forced by the Greenland ice core palaeo-temperature
      records (GRIP, NGRIP) produce the highest variability in modelled
      sea-level relevant ice
      volume throughout the last glacial cycle. In contrast, simulations
      driven by oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
      palaeo-temperature records generally result in lower ice volume
      variability throughout the simulation length, resulting in lower
      modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
      The NGRIP climate forcing is the only one that results in a~larger ice
      volume during MIS~4 than during MIS~2.

      While simulations driven by the GRIP and the two Antarctic
      palaeo-temperature records attain a~last sea-level relevant ice volume
      maximum between
      19.1 and 16.9\,\unit{ka}, those informed by the NGRIP and the two
      oceanic palaeo-temperature records attain their maximum ice volumes
      thousands of years earlier. Moreover, the ODP~1012 run yields a~rapid
      deglaciation of the modelled grounded ice extent prior to
      17\,\unit{ka}. The ODP~1020
      simulation predicts an early maximum in ice volume at 20.4\,\unit{ka},
      followed by slower deglaciation than modelled using the other
      palaeo-temperature records. Finally, whereas model runs forced by the
      Antarctic palaeo-temperature records result in a~rapid and
      uninterrupted deglaciation after the LGM, the simulation driven by the
      GRIP palaeo-temperature record also results in a~rapid deglaciation
      but in three steps, separated by two periods of ice sheet regrowth
      (Fig.~\ref{fig:lr_ts}).

@@

      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled sea-level relevant ice volume
      minima spread over
      a~wide range between 1.54 and 2.81\,\unit{m}\,s.l.e.
      (Table~\ref{tab:extrema}).

@@

      As a~direct result of the different scaling factors applied
      (Table~\ref{tab:sens_params}), the resulting simulations show very
      little difference in the modelled glaciated areas corresponding to
      the maximum stage during MIS~2, but also during MIS~4
      (Table~\ref{tab:sens_extrema}). However, such can not be said of the
      modelled glaciated area corresponding to minimum stage during
      MIS~3. In fact, the extent of the remnant ice cap which persists over
      the Skeena Mountains during this stage shows a~significant sensitivity
      to ice rheology of 25\,\unit{\%}, and an even more important
      sensitivity to basal sliding parameters of 57\,\unit{\%}
      (Table~\ref{tab:sens_extrema}).

      The modelled sea-level potentials show more variability than
      the modelled glaciated areas (Table~\ref{tab:sens_extrema},
      Fig.~\ref{fig:sens_ts}). As one could expect, softer ice and weaker
      till both result in a~thinner ice sheet, while harder ice and stronger
      till result in a~thicker ice sheet. For instance, peak sea-level
      relevant ice volume
      during the MIS~2 (LGM) varies by 25\,\unit{\%} between the two
      parametrisations of ice rheology used, and by 14\,\unit{\%} between
      the two parametrisations of basal sliding used. The differences in
      ice volume are greatest during the MIS~3
      (Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the
      areal and thickness contributions add up.

@@

      The exact timing of modelled MIS~2 maximum ice volume depends strongly
      on the choice of applied palaeo-temperature record, which allows for
      a~more in-depth comparison with geological evidence for the timing of
      maximum Cordilleran ice sheet extent. In the Puget Lowland
      (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
      ice sheet margin has been constrained by radiocarbon dating on wood
      between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
      \citep{Porter.Swanson.1998}. These dates are consistent with
      radiocarbon dates from the offshore sedimentary record, which reveals
      an increase of glaciomarine sedimentation between 19.5 and
      16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008,
      Taylor.etal.2014}. Radiocarbon dating of the northern Cordilleran ice
      sheet margin is much less constrained but straddles presented
      constraints from the southern margin. However, cosmogenic exposure
      dating places the timing of maximum northern ice sheet margin extent
      during the McConnell
      glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A~sharp transition in
      the sediment record of the Gulf of Alaska indicates a~retreat of
      regional outlet glaciers onto land at
      14.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Davies.etal.2011}.

@@

      Such deviation from the geological inferences could reflect the fact
      the NARR has difficulties with simulating orographic processes in some
      areas of steep topography \citep{Jarosch.etal.2012}. In a~previous
      study \citep{Seguinot.etal.2014}, we have evaluated the performance of
      NARR in forcing constant-climate simulations of the Cordilleran ice
      sheet against that of an observation-based data set
      \citep[WorldClim,][]{Hijmans.etal.2005}. Indeed, the use of NARR in
      these simulations produced slightly different patterns of glaciation
      relative to WorldClim, including more extensive ice cover on the
      Columbia and Rocky mountains \citep[Figs.~6--7]{Seguinot.etal.2014}.
      Our simulations have shown that these differences are mainly caused by
      disparities in the precipitation fields of the two data sets
      \citep[Figs.13--14]{Seguinot.etal.2014}. Over the southern part of our
      model domain, it was shown that the implementation of temperature and
      precipitation downscaling can address these limitations
      \citep[e.g.,][]{Jarosch.etal.2012}. However, extending this
      downscaling method to the entire model domain used in our study could
      be challenging, because the northern part of the model domain is
      characterized by stronger precipitation gradients (Fig.~\ref{fig:atm})
      and includes much fewer weather stations than the south
      \citep{Hijmans.etal.2005}.

      The easterly position of the modelled ice divide and eastern ice sheet
      margin may also reflect the fact that our model does not include
      feedback mechanisms between ice sheet topography and the regional
      climate since it assumes fixed modern-day spatial patterns for
      temperature and precipitation. In fact, it is reasonable to think
      that the cooling was greater inland than near the coast, prohibiting
      melt at the eastern margin. However, our simulations already produce
      an excess of ice inland. Including such a~temperature continentality
      gradient in the model while keeping the precipitation pattern constant
      would thus cause an even greater mismatch between the model results
      and the geologically reconstructed ice margins during the LGM.

>>>>

      Despite large differences in the input climate forcing
      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
      features that can be observed across the range of forcing data used.
      In all simulations, modelled sea-level relevant ice volumes remain
      relatively low during
      most of the glacial cycle, except during two major glacial events
      which occur between 62.2 and 56.9\,\unit{ka} during MIS~4, and between
      23.2 and 16.9\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
      panel). An ice volume minimum is consistently
      reached between 55.9 and
      42.9\,\unit{ka} during MIS~3. However, the magnitude and precise
      timing of these three events depend significantly on the choice of
      proxy record used to derive a~time-dependent climate forcing
      (Table~\ref{tab:extrema}).

      Simulations forced by the Greenland ice core palaeo-temperature
      records (GRIP, NGRIP) produce the highest variability in modelled ice
      volume throughout the last glacial cycle. In contrast, simulations
      driven by oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
      palaeo-temperature records generally result in lower ice volume
      variability throughout the simulation length, resulting in lower
      modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
      The NGRIP climate forcing is the only one that results in a~larger ice
      volume during MIS~4 than during MIS~2.

      While simulations driven by the GRIP and the two Antarctic
      palaeo-temperature records attain a~last ice volume maximum between
      19.1 and 16.9\,\unit{ka}, those informed by the NGRIP and the two
      oceanic palaeo-temperature records attain their maximum ice volumes
      thousands of years earlier. Moreover, the ODP~1012 run yields a~rapid
      deglaciation of the modelled grounded ice extent prior to
      17\,\unit{ka}. The ODP~1020
      simulation predicts an early maximum in ice volume at 20.4\,\unit{ka},
      followed by slower deglaciation than modelled using the other
      palaeo-temperature records. Finally, whereas model runs forced by the
      Antarctic palaeo-temperature records result in a~rapid and
      uninterrupted deglaciation after the LGM, the simulation driven by the
      GRIP palaeo-temperature record also results in a~rapid deglaciation
      but in three steps, separated by two periods of ice sheet regrowth
      (Fig.~\ref{fig:lr_ts}).

@@

      During MIS~3 ice volume minimum reconstructions, a~central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled sea-level relevant ice volume
      minima spread over
      a~wide range between 1.54 and 2.81\,\unit{m}\,s.l.e.
      (Table~\ref{tab:extrema}).

@@

      As a~direct result of the different scaling factors applied
      (Table~\ref{tab:sens_params}), the resulting simulations show very
      little difference in the modelled glaciated areas corresponding to
      the maximum stage during MIS~2, but also during MIS~4
      (Table~\ref{tab:sens_extrema}). However, this cannot be said of the
      modelled glaciated area corresponding to the minimum stage during
      MIS~3. In fact, the extent of the remnant ice cap which persists over
      the Skeena Mountains during this stage shows a~significant sensitivity
      to ice rheology of 25\,\unit{\%}, and an even higher
      sensitivity to basal sliding parameters of 57\,\unit{\%}
      (Table~\ref{tab:sens_extrema}).

      The modelled sea-level potentials show a stronger variability than
      the modelled glaciated areas (Table~\ref{tab:sens_extrema},
      Fig.~\ref{fig:sens_ts}). As one could expect, softer ice and weaker
      till both result in a~thinner ice sheet, while harder ice and stronger
      till result in a~thicker ice sheet. For instance, the modelled peak
      sea-level relevant ice volume
      during MIS~2 (LGM) varies by 25\,\unit{\%} between the two
      parametrisations of ice rheology used, and by 14\,\unit{\%} between
      the two parametrisations of basal sliding used. The differences in
      ice volume are greatest during MIS~3
      (Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the
      areal and thickness contributions add up.

@@

      The exact timing of modelled MIS~2 maximum ice volume depends strongly
      on the choice of applied palaeo-temperature record, which allows for
      a~more in-depth comparison with geological evidence for the timing of
      the maximum Cordilleran ice sheet extent. In the Puget Lowland
      (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
      ice sheet margin has been constrained by radiocarbon dating on wood
      between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
      \citep{Porter.Swanson.1998}. These dates are consistent with
      radiocarbon dates from the offshore sedimentary record, which reveals
      an increase of glaciomarine sedimentation between 19.5 and
      16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008,
      Taylor.etal.2014}. Radiocarbon dating of the northern Cordilleran ice
      sheet margin is much less constrained but straddles presented
      constraints from the southern margin. However, cosmogenic exposure
      dating places the timing of the maximum northern ice sheet margin
      extent during the McConnell
      glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A~sharp transition in
      the sediment record of the Gulf of Alaska indicates a~retreat of
      regional outlet glaciers onto land at
      14.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Davies.etal.2011}.

@@

      Such deviation from the geological inferences could reflect the fact
      that the NARR has difficulties with simulating orographic processes in
      some areas of steep topography \citep{Jarosch.etal.2012}. In our
      previous study \citep{Seguinot.etal.2014}, we have evaluated an
      applicability of the climate forcing bilinearly interpolated from the
      NARR to constant-climate simulations of the Cordilleran ice sheet
      during the LGM against that of an observation-based data set
      \citep[WorldClim,][]{Hijmans.etal.2005}. Indeed, the use of NARR in
      these simulations produced slightly different patterns of glaciation
      relative to WorldClim, including a more extensive ice cover on the
      Columbia and Rocky mountains \citep[Figs.~6--7]{Seguinot.etal.2014}.
      Our simulations have shown that these differences are mainly caused by
      disparities between the precipitation fields of the two data sets
      \citep[Figs.13--14]{Seguinot.etal.2014}. It has been shown that over
      the southern part of our model domain that temperature and
      precipitation downscaling can potentially address these limitations
      \citep[e.g.,][]{Jarosch.etal.2012}. However, extending this
      downscaling method to the entire model domain used in our study is
      challenging, because the northern part of the model domain is
      characterized by stronger precipitation gradients (Fig.~\ref{fig:atm})
      and fewer weather stations than the southern part where the previous
      analysis has been performed \citep{Hijmans.etal.2005}.

      Since the model does not include feedback mechanisms between the ice
      sheet topography and the regional climate, the modelled easterly
      positions of the ice divide and eastern ice sheet margin may also be
      sensitive to the assumption of fixed modern-day spatial patterns of
      air temperature and precipitation. In fact, it is reasonable
      to think that the cooling during the last glacial cycle was greater
      inland than near the coast, prohibiting
      melt at the eastern margin. However, our simulations already produce
      an excess of ice inland. Including such a~temperature continentality
      gradient in the model while keeping the precipitation pattern constant
      would thus cause an even greater mismatch between the model results
      and the geologically reconstructed ice margins during the LGM.
